\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage[colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black, urlcolor=blue!50!black]{hyperref}

\newcommand{\F}{\mathbb{F}}
\newcommand{\K}{K}            % F_{2^64}: addresses, counters, committed lanes
\newcommand{\E}{E}            % F_{2^128} = K[y]/(y^2+x y+1): words and challenges
\newcommand{\pkd}{\mathrm{pkd}}
\newcommand{\eq}{\operatorname{eq}}
\newcommand{\cube}[1]{\{0,1\}^{#1}}
\newcommand{\cin}{\mathit{cin}}
\newcommand{\oA}{\widehat A}
\newcommand{\oB}{\widehat B}
\newcommand{\oC}{\widehat C}
\newcommand{\ocout}{\widehat{c}_{\mathrm{out}}}
\newcommand{\ocin}{\widehat{c}_{\mathrm{in}}}
\newcommand{\oQ}{\widehat Q}
\newcommand{\II}{\widehat I}
\newcommand{\D}{D}

% VM macros
\newcommand{\gen}{\mathsf{g}}
\newcommand{\loc}[1]{\mem[\fp\cdot #1]}
\newcommand{\pc}{\mathbf{pc}}
\newcommand{\fp}{\mathbf{fp}}
\newcommand{\mem}{\mathbf{m}}
\newcommand{\addr}{\mathit{addr}}
\newcommand{\val}{\mathit{val}}
\newcommand{\cnt}{\mathit{count}}
\newcommand{\push}{\textsc{push}}
\newcommand{\pull}{\textsc{pull}}
\newcommand{\tup}[1]{\langle #1 \rangle}
\newcommand{\dsep}[1]{\mathsf{#1}}
\newcommand{\opc}[1]{\mathsf{op}_{\mathsf{#1}}}
\newcommand{\srcsel}[1]{\mathsf{src}(#1)}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO:} #1\textbf{]}}}

\title{\bf leanVM-b}
\author{}
\date{}

\begin{document}
\maketitle


\tableofcontents
\newpage

\section{VM specification}\label{sec:vm}

\paragraph{Field.}
\[
  \K=\F_{2^{64}}=\F_2[x]/(x^{64}+x^4+x^3+x+1),
  \qquad
  \E=\K[y]/(y^2+x y+1),
  \qquad |\E|=2^{128}.
\]
A \emph{memory word}, including a cell value, immediate, or hash value, is an element $v=v_0+v_1y\in\E$ with $v_0,v_1\in\K$. Each address, program counter $\pc$, frame pointer $\fp$, and read counter is in $\K$. Addition is bitwise \textsc{xor}; multiplication is in $\K$ or $\E$. Protocol randomness is sampled from $\E$ (\S\ref{sec:prelim}). Each committed column (\S\ref{sec:rs}) is $\K$-valued, so a data word occupies two columns.

\paragraph{Addresses as powers of a generator.} Fix a generator $\gen$ of the multiplicative group $\K^{\times}$, of order $2^{64}-1$. Memory addresses and bytecode addresses (the program counter) are powers of $\gen$: the $i$-th address is the field element $\gen^{\,i}$. Its successor is one field multiplication: $\gen^{\,i+1}\;=\;\gen\cdot\gen^{\,i}$.

\paragraph{Memory.} Memory is read-only (write-once): an array of $2^{h}$ cells, each holding one $128$-bit word (an $\E$-element), addressed by the $2^{h}$ powers $\gen^{0},\gen^{1},\dots,\gen^{2^{h}-1}$ (the addresses are $\K$-elements). The size exponent $h$ is fixed per execution, with $16\le h\le 32$. Each cell is set once; every access thereafter reads it.

\paragraph{Program.} The program is a fixed, public sequence of instructions, addressed like memory by powers of $\gen$: the program counter $\gen^{\,i}$ selects the $i$-th instruction. An instruction is an opcode together with its operands, described below.

\paragraph{Registers.} The machine state is two registers, the program counter $\pc$ and the frame pointer $\fp$, each a power of $\gen$. Every instruction except \texttt{JUMP} advances the state by $\pc\gets\gen\cdot\pc$ (fetching the next instruction) and leaves $\fp$ unchanged; \texttt{JUMP} may set both.

\paragraph{Operands and addressing.} An operand is either an \emph{immediate}, a literal field element used as given, or an \emph{fp-relative reference}: an offset $j$ into the current frame, encoded as the $\gen$-power $o=\gen^{\,j}$. With the frame based at $\fp=\gen^{\mathrm{base}}$, the reference names the cell
\[
  \mem[\fp\cdot o],\qquad \fp\cdot o\;=\;\gen^{\mathrm{base}}\cdot\gen^{\,j}\;=\;\gen^{\,\mathrm{base}+j}.
\]
Since addresses are exponents of $\gen$, the frame is the contiguous block $\gen^{\mathrm{base}},\gen^{\mathrm{base}+1},\dots$, and adding the offset $j$ to the base exponent is, in the field, a single multiplication $\fp\cdot o$.

\paragraph{Instruction set.} The machine implements seven instructions. Below, $o,o_A,\dots$ are fp-relative references and $k$ an immediate operand.
\begin{center}
\begin{tabular}{lll}
\hline
Instruction & Operands & Semantics\\
\hline
\texttt{XOR}           & $[o_A,o_B,o_C]$ & $\loc{o_C}=\loc{o_A}+\loc{o_B}$ \quad($128$-bit field addition, \textsc{xor})\\
\texttt{MUL\_NATIVE}   & $[o_A,o_B,o_C]$ & $\loc{o_C}=\loc{o_A}\cdot\loc{o_B}$ in $\E$\\
\texttt{SET\_CONSTANT} & $[o,k]$         & $\loc{o}=k$ \quad($k$ a $128$-bit immediate)\\
\texttt{DEREF}         & $[\alpha,\beta,\gamma;\,s]$ & $\mem[\loc{\alpha}\cdot\beta]=\srcsel{s}$\\
\texttt{JUMP}          & $[o_c,o_d,o_f]$ & conditional jump (below)\\
\texttt{PACK64X2}      & $[o_A,o_B,o_C]$ & $\loc{o_C}=a_0+b_0y$ for $\loc{o_A}=a_0,\loc{o_B}=b_0\in\K$\\
\texttt{BLAKE3}        & $[o_0,o_1,o_2,o_3,o_C]$ & $c=\mathrm{BLAKE3}(a,b)$ on independently-addressed single-cell chunks (below)\\
\hline
\end{tabular}
\end{center}

\texttt{PACK64X2} groups two $64$-bit cells into one canonical $128$-bit cell. Its memory interactions read the sources as $(a_0,0)$ and $(b_0,0)$ and the destination as $(a_0,b_0)$. Bus balance therefore proves that both source words lie in $\K$; this is not merely a witness-generation check. The instruction is useful when a program receives serialized $\K$-limbs through an $\E$-valued interface.

\texttt{BLAKE3} compresses $64$ bytes to $32$. Each $128$-bit chunk occupies one cell. References $o_0,\dots,o_3$ independently address the four input cells, with $\loc{o_i}=\fp\cdot o_i$. The operands are $a=\loc{o_0}\Vert\loc{o_1}$ and $b=\loc{o_2}\Vert\loc{o_3}$. The input cells need not be adjacent. The digest occupies the consecutive cells $\loc{o_C}$ and $\loc{\gen\cdot o_C}$. Each row therefore accesses six cells. The BLAKE3 table (\S\ref{sec:tab-blake3}) connects them to the memory, state, and bytecode interactions. Flock~\cite{flock} proves $c=\mathrm{BLAKE3}(a,b)$ with an R1CS proof of the compression circuit.

\texttt{DEREF} stores, at the dereferenced address $\loc{\alpha}\cdot\beta$, a value chosen by a \emph{store mode} $s\in\{\texttt{deref\_cell}[\gamma],\texttt{deref\_pc},\texttt{deref\_fp}\}$:
\[
  \srcsel{s}\;=\;
  \begin{cases}
    \loc{\gamma} & s=\texttt{deref\_cell}[\gamma] \quad(\text{the local cell at offset }\gamma),\\
    \gen^{2}\cdot\pc & s=\texttt{deref\_pc} \quad(\text{a return address: the program counter advanced by two}),\\
    \fp & s=\texttt{deref\_fp} \quad(\text{the current frame pointer}).
  \end{cases}
\]
The \texttt{deref\_pc} and \texttt{deref\_fp} modes let a caller save its return address and frame so a callee can later return and restore them, the basis for function calls. The skip of two is the call's return target: place the \texttt{DEREF} that saves $\gen^{2}\cdot\pc$ immediately before the \texttt{JUMP} that transfers control, and $\gen^{2}\cdot\pc$ names the instruction just after that \texttt{JUMP}.

\texttt{JUMP} reads a condition $c=\loc{o_c}$, an arbitrary $128$-bit word, and branches on whether it is zero. When $c\neq0$ it transfers control, $\pc\gets\loc{o_d}$ and $\fp\gets\loc{o_f}$; when $c=0$ it falls through, $\pc\gets\gen\cdot\pc$ and $\fp\gets\fp$.

\paragraph{Words used as addresses.} A memory word is $128$-bit, while $\pc$, $\fp$, and addresses are in $\K$. Three reads produce addresses or registers: the \texttt{DEREF} pointer $\loc{\alpha}$, the \texttt{JUMP} destination $\loc{o_d}$, and the \texttt{JUMP} frame $\loc{o_f}$. Each must have high lane zero. The address and selection constraints (\S\ref{sec:tab-deref}, \S\ref{sec:tab-jump}) equate these words, possibly multiplied by a $\K$-operand, to a $\K$ column.


\iffalse % Deferred future instructions (u32, BLAKE3): commented out for now.
\noindent The remaining instructions are listed to fix the eventual ISA but are deferred to a later version.
\begin{center}
\begin{tabular}{lll}
\hline
Instruction (later) & Operands & Semantics\\
\hline
\texttt{u32\_ADD}    & $[o_A,o_B,o_C]$     & integer add mod $2^{32}$\\
\texttt{IS\_NONZERO} & $[o_x,o_b]$         & $\loc{o_b}=[\,\loc{o_x}\neq 0\,]$\\
\texttt{u32\_MUL}    & $[o_A,o_B,o_C,o_D]$ & integer multiply, double word: $a\cdot b=c+2^{32}d$\\
\hline
\end{tabular}
\end{center}

\noindent These operate on $32$-bit integers (\texttt{u32}) rather than native field elements, and so require integer addition with carries, which the field arithmetic does not provide. Their full semantics are fixed later. \todo{specify the \texttt{u32\_MUL} semantics later.}
\fi % end deferred future instructions

\section{Proving primitives}\label{sec:prelim}

Every physical committed column is $\K$-valued (\S\ref{sec:vm}); every challenge, sumcheck, zerocheck, GKR value, and evaluation claim is $\E$-valued, where
\[
  \K=\F_2[x]/(x^{64}+x^4+x^3+x+1),
  \qquad
  \E=\K[y]/(y^2+x y+1)\cong\F_{2^{128}}.
\]
Thus $|\E|=2^{128}$. The embedding and mixed product are
\[
  \K\hookrightarrow\E,\quad e\mapsto e+0\cdot y,
  \qquad
  k(e_0+e_1y)=ke_0+(ke_1)y.
\]
Committed columns and their truth tables are $\K$-valued; extensions are evaluated at $\E$-points; protocol randomness is sampled from $\E$. A $128$-bit \emph{value} column is represented by $V_0,V_1:\cube\ell\to\K$ with $V=V_0+V_1y$. Address, register, and counter columns are single $\K$-columns.

\textbf{Multilinear extension (MLE).} A function $b:\cube{\ell}\to\K$ (a committed column) or $b:\cube{\ell}\to\E$ (a product-tree layer of \S\ref{sec:gkr}) has a unique multilinear extension $\widetilde b:\E^{\ell}\to\E$, the polynomial of degree at most $1$ in each variable that agrees with $b$ on the boolean hypercube. The equality polynomial $\eq(r,X)=\prod_i(1+r_i+X_i)$ is the multilinear that is $1$ at $X=r$ and $0$ at every other Boolean point.

\textbf{Tables and columns.} A \emph{table} is a group of columns of the same size. Each column is \emph{committed} or \emph{virtual}: a virtual column is a $\K$-linear function of the committed columns of its table, carried with no commitment of its own. Such a map, extended $\E$-linearly, commutes with the multilinear extension, $\widetilde{L(x)}(\zeta)=L\bigl(\widetilde x(\zeta)\bigr)$, so one opening of the committed columns evaluates the virtual one.

\textbf{Sumcheck.} Sumcheck reduces a claim $\sum_{x\in\cube\ell}P(x)=s$ to a single evaluation $P(\rho)$ at a random $\rho\in\E^{\ell}$, over $\ell$ rounds; a false claim passes with probability at most $\deg P\cdot\ell/|\E|$.

\textbf{Zerocheck.} To prove $P$ vanishes on $\cube\ell$, the verifier samples $r\in\E^{\ell}$ and runs sumcheck on $\sum_{x}P(x)\,\eq(r,x)=0$.

\textbf{Inner-product PCS.} An \emph{inner-product commitment scheme} lets a prover commit to a multilinear $g:\cube{M}\to\K$ and later prove a weighted sum $\sum_{w\in\cube M} W(w)\,\widehat g(w)=c$ against any weight $W:\cube{M}\to\E$ the verifier can evaluate: the commitment is over $\K$, the opening over $\E$. A point evaluation $\widehat g(\mathbf r)$ at $\mathbf r\in\E^{M}$ is the special case $W=\eq(\mathbf r,\cdot)$. WHIR and BaseFold/Ligerito are such schemes; we use one (\S\ref{sec:rs-dense}).

\section{Committing the witness}\label{sec:rs}

\subsection{Stacking}\label{sec:stacking}

No column is committed on its own. The columns are $\K$-valued multilinears over boolean hypercubes of various sizes, concatenated into one witness $q$; one region of the stack is flock's packed Boolean witness $q_\pkd$, $64$ bits per $\K$-word (\S\ref{sec:tab-blake3}). Order the $T$ columns $\widehat P_1,\dots,\widehat P_T$ by size, $\kappa_1\ge\dots\ge\kappa_T$ variables, and lay their truth tables end to end,
\[
  q \;=\; P_1 \,\Vert\, P_2 \,\Vert\,\dots\,\Vert\, P_T \,\Vert\, 0,
\]
zero-padded to length $2^{M}$, with $M=\bigl\lceil\log_2\textstyle\sum_i 2^{\kappa_i}\bigr\rceil$. Placed largest first, column $i$ starts at an offset that is a multiple of $2^{\kappa_i}$; writing $\mathsf{sel}_i\in\cube{M-\kappa_i}$ for the high bits of that offset,
\[
  \widehat P_i(z)\;=\;\widehat q(z,\mathsf{sel}_i)\qquad(z\in\cube{\kappa_i}).
\]
An evaluation claim $\widehat P_i(\zeta)=c$ is therefore a claim on the stack,
\[
  \widehat q(\zeta,\mathsf{sel}_i)\;=\;\sum_{w\in\cube{M}}\eq\bigl((\zeta,\mathsf{sel}_i),w\bigr)\,\widehat q(w)\;=\;c,
\]
a weighted sum over $\widehat q$ with weight $W(w)=\eq((\zeta,\mathsf{sel}_i),w)$.

\paragraph{Batching.} Every claim the protocol raises against a column is such a weighted sum $\sum_w W_j(w)\,\widehat q(w)=c_j$, with $\E$-valued weight and target. The verifier folds the $J$ of them with independent random coefficients $\lambda_j\in\E$, one per claim, each sampled after the claimed values are bound, into one weight $W_\lambda=\sum_j\lambda_j W_j$ and target $C_\lambda=\sum_j\lambda_j c_j$, and the inner-product PCS (\S\ref{sec:rs-dense}) proves
\[
  \sum_{w\in\cube{M}} W_\lambda(w)\,\widehat q(w)=C_\lambda
\]

\subsection{Inner-product PCS}\label{sec:rs-dense}

It is Ligerito~\cite{cryptoeprint:2025/1187} in the \emph{Secure} profile: rate $1/2$, unique-decoding regime (list size $1$, no out-of-domain samples), $120$-bit round-by-round soundness. The witness is Reed--Solomon encoded over $\K$; level-0 symbols are $8$ bytes and $2^6$ interleaved symbols form each $512$-byte Merkle leaf. Openings are over $\E$: every row-batching and folding challenge and the opening sumcheck are sampled from $\E$, so the level-0 fold pairs $\K$-symbols with $\E$-challenges and all deeper levels are $\E$-valued.

\section{M3 model}\label{sec:m3}

The arithmetization follows the M3 model (multi-multiset matching)~\cite{m3}. The prover chooses each table's size, rows are unordered, and an instance is accepted exactly when every table's constraints hold and the bus balances.

\paragraph{Constraints.} A constraint is a polynomial in the columns that must vanish on every row, including the padding rows.

\paragraph{The bus.} All interactions share a single \emph{bus}, a channel carrying tuples of $m$ field elements. A table flushes by wiring its columns into a tuple and either \emph{pushing} it onto the bus or \emph{pulling} it off, once per row. A few fixed \emph{boundary} tuples, belonging to no table, are pushed or pulled directly. The width $m$ is that of the widest tuple, nine (the bytecode tuple: separator, $\pc$, count, opcode, three operands, and two \texttt{DEREF} flags); shorter tuples are zero-padded.

\paragraph{Domain separation.} The first coordinate of every tuple is a \emph{domain separator}, a fixed constant naming its interaction: $\dsep{ST}$ for state, $\dsep{MEM}$ for memory, $\dsep{BC}$ for bytecode. Since everything shares the one bus, the separators are what keep the three kinds of tuple from cancelling against one another: a memory tuple can never meet a bytecode tuple. The interactions below are three uses of this one channel, told apart by this coordinate.

\paragraph{Balance.} The bus balances when its pushed tuples and its pulled tuples form the same multiset: the same tuples with the same multiplicities, in any order. This is the instance's only global check, proven by grand product argument (\S\ref{sec:gp}).

\paragraph{Per-instruction tables.} Each instruction has its own table, with columns $\pc$, $\fp$, $\mathit{next\_pc}$, $\mathit{next\_fp}$, the operands, the operand addresses, and the read values. For every instruction except \texttt{JUMP}, the successor is virtual: $\mathit{next\_fp}=\fp$ is a copy of $\fp$, and $\mathit{next\_pc}=\gen\cdot\pc$ is the free increment of \S\ref{sec:vm}. Each row makes the same kinds of flush: on the state interaction (\S\ref{sec:state}) it pulls $\tup{\dsep{ST},\pc,\fp}$ and pushes $\tup{\dsep{ST},\mathit{next\_pc},\mathit{next\_fp}}$; it makes one memory flush per cell read (\S\ref{sec:memchan}) and one bytecode flush for the instruction at $\pc$ (\S\ref{sec:bytecode}). The concrete tables are in \S\ref{sec:tables}.

\subsection{Proving the local constraints}\label{sec:air}

A proof is valid when every table satisfies its local constraints and the bus balances; this subsection handles one table's constraints, \S\ref{sec:gp} the balance. Each table's constraints are batched into a single zerocheck of their own, with no batching across tables.

Fix a table with $2^{\tau}$ rows on the cube $\cube\tau$, and let $C_1,\dots,C_s$ be its constraints: polynomials with $\K$-coefficients in the row's column values, each of degree $\le d$ and required to vanish on every row ($d=2$ throughout).

\paragraph{Batch the constraints.} The verifier samples $\eta\in\E$ and forms the combined constraint
\[
  C\;=\;\sum_{t=1}^{s}\eta^{\,t-1}\,C_t .
\]
If some $C_t$ is nonzero on a row, so is $C$ there, except with probability $(s-1)/|\E|$ over $\eta$. It thus suffices to show $C$ vanishes on $\cube\tau$.

\paragraph{Zerocheck.} The verifier samples $r\in\E^{\tau}$ and the parties run the zerocheck of \S\ref{sec:prelim}, the sumcheck
\[
  \sum_{x\in\cube\tau} C(x)\,\eq(r,x)\;=\;0 .
\]
The implementation factors the known $\eq$ term out of each round. It sends the degree-$d$ cofactor at the three points $0,1,\gen$ (so three $\E$-elements for $d=2$), and the verifier restores the linear equality factor. After $\tau$ rounds the claim is $C(\rho)\,\eq(r,\rho)$ at a random $\rho\in\E^{\tau}$.

\paragraph{Discharge.} To check the final value $C(\rho)\,\eq(r,\rho)$, the verifier computes $\eq(r,\rho)$ itself and assembles $C(\rho)=\sum_t\eta^{\,t-1}C_t(\rho)$ from the column values at $\rho$: the prover sends one value per committed column, each virtual column read off its source for free. These prover-sent values are checked later against the commitments by the PCS  (\S\ref{sec:stacking}).

\subsection{Balancing the bus: grand product}\label{sec:gp}

The bus balances when its pushed and pulled tuples form the same multiset (\S\ref{sec:m3}). Two random challenges, $\alpha$ then $\gamma$, reduce this to a single product.

\emph{Fingerprint.} The verifier samples $\alpha\in\E$ and maps each width-$m$ tuple $\sigma=(\sigma_0,\dots,\sigma_{m-1})$ to a single $\E$-element,
\[
\pi_\alpha(\sigma)\;=\;\sum_{i=0}^{m-1}\alpha^{i}\,\sigma_i ,
\]
each coordinate $\sigma_i$ a committed column value, a virtual column value, or a public constant, all $\K$-elements, so each summand is a mixed $\K\times\E$ product.

\emph{Product check.} The verifier samples $\gamma\in\E$ and the prover proves
\begin{equation}
\prod_{\sigma\ \text{pushed}}\bigl(\gamma-\pi_\alpha(\sigma)\bigr)
\;=\;
\prod_{\sigma\ \text{pulled}}\bigl(\gamma-\pi_\alpha(\sigma)\bigr).
\label{eq:gp}
\end{equation}
\emph{Completeness and soundness.} If the pushed and pulled multisets agree, the two products in \eqref{eq:gp} are equal for every $\alpha,\gamma$. Otherwise their difference is a nonzero polynomial in $(\alpha,\gamma)$ of total degree at most $2^\mu(m-1)$. The separate count product contributes at most $2^{\mu_c}/|\E|$, where $\mu_c\leq\mu$. Hence the total error is at most
\[
  \frac{2^\mu(m-1)+2^{\mu_c}}{2^{128}}.
\]
Here $m=9$, so the prover grinds a Fiat--Shamir nonce for
\[
  \max(0,\,120+\mu+4-128)
\]
bits before sampling $\alpha$ and $\gamma$.

Each side's product is proven by a GKR pass over a product tree (\S\ref{sec:gkr}).

\paragraph{Padding.} Every table is padded past its real rows to a power of two. Each padding row is a fixed default: every column zero but the read counts, set to $\gen^{0}=1$. Its memory and bytecode flushes are then default tuples that do not self-cancel, so the prover announces each table's real row count and the verifier divides the surplus out of \eqref{eq:gp}: a default tuple $d$ in surplus $\delta_d$ on a side divides that side by $(\gamma-\pi_\alpha(d))^{\delta_d}$.


\subsection{Proving the grand product (GKR)}\label{sec:gkr}

Take one side of \eqref{eq:gp}. Its tuples give the leaves $v_k=\gamma-\pi_\alpha(\sigma_k)$, padded with $1$s to a power of two, $2^\mu$ in all, and the product $P=\prod_k v_k$ is the root of a binary tree: layer $0$ is the leaves, layer $i$ has $2^{\mu-i}$ nodes, each the product of its two children, and layer $\mu$ is $P$. Let $\widetilde V_i$ be the MLE of the layer-$i$ values $V_i:\cube{\mu-i}\to\E$ (the leaves are $\E$-valued, being $\gamma$ minus a fingerprint). Adjacent layers obey $V_i(x)=V_{i-1}(0,x)\,V_{i-1}(1,x)$ on the cube, so for any $r$,
\[
\widetilde V_i(r)=\sum_{x\in\cube{\mu-i}}\eq(r,x)\,\widetilde V_{i-1}(0,x)\,\widetilde V_{i-1}(1,x).
\]
GKR walks from the root to the leaves, reducing one claim per layer. Starting from a claim $\widetilde V_i(r)$:
\begin{enumerate}
\item a $(\mu-i)$-round sumcheck of the identity above reduces it to the two values $\widetilde V_{i-1}(0,\rho)$, $\widetilde V_{i-1}(1,\rho)$ at a random $\rho$. The round univariate has degree $3$, but its $\eq(r,\cdot)$ factor is linear in the bound variable and known to the verifier, so the prover sends only the degree-$2$ cofactor (the product part) as $3$ evaluations and the verifier multiplies the $\eq$ factor back~\cite{cryptoeprint:2024/108}, making each round's work degree $2$ rather than $3$;
\item the prover sends the degree-$1$ line $t\mapsto\widetilde V_{i-1}(t,\rho)$; the verifier checks its endpoints against those two values and evaluates it at a random $c$, leaving one claim $\widetilde V_{i-1}(c,\rho)$.
\end{enumerate}
After $\mu$ layers a single leaf claim $\widetilde V_0(\zeta)$ remains, which \S\ref{sec:leafstack} decomposes into evaluation claims on the committed columns. The bus is accepted when the two sides of \eqref{eq:gp} give equal roots and each leaf claim matches the commitments. Prover work is $O(2^\mu)$, and the proof is $O(\mu^2)$ field elements.

\paragraph{Batching the three trees.} The bus needs three grand products: push, pull, and counts. They are batched in a single GKR pass (verifier-friendliness, good for recursion). Push and pull have equal depth $\mu$ (their blocks come in matched pairs); the count tree, of depth $\mu_c\le\mu$, is padded with $1$-leaves to $2^{\mu}$.

The verifier samples a combiner $\lambda$, and each layer runs a single sumcheck on the $\lambda$-combined layer identity
\[
\sum_{t}\lambda^{t}\,\widetilde V^{t}_i(r)\;=\;\sum_{x\in\cube{\mu-i}}\eq(r,x)\,\sum_{t}\lambda^{t}\,\widetilde V^{t}_{i-1}(0,x)\,\widetilde V^{t}_{i-1}(1,x),\qquad t\in\{0,1,2\},
\]

\subsection{Stacking the leaves}\label{sec:leafstack}

GKR (\S\ref{sec:gkr}) reduces one side of \eqref{eq:gp} to a single evaluation $\widetilde V_0(\zeta)$ of its \emph{leaves}, the product's factors $\gamma-\pi_\alpha(\sigma)$, one per bus tuple $\sigma$. We settle this evaluation against the committed columns.

\paragraph{Flush blocks.} The leaves split into \emph{blocks}, one per flush rule and one per boundary set. A block of a $2^{\kappa}$-row table holds $2^{\kappa}$ leaves; the row-$z$ leaf is $\gamma-\sum_{i=0}^{m-1}\alpha^{i}\,c_{i}(z)$, with $c_{i}(z)$ coordinate $i$ of the row's tuple. Each $c_i$ is public, a committed column, or a virtual column.

\paragraph{Layout.} Order the blocks largest first at aligned offsets, padding to $2^{\mu}$ leaves with $1$s, which leave the product unchanged. This is the stacking of \S\ref{sec:stacking} applied to blocks: block $b$, of $2^{\kappa_b}$ leaves, occupies the subcube $\{(z,\mathsf{sel}_b):z\in\cube{\kappa_b}\}$ for a fixed selector $\mathsf{sel}_b$.

\paragraph{Decomposition.} Split $\zeta=(\zeta^{b}_{\mathrm{lo}},\zeta^{b}_{\mathrm{hi}})$ at block $b$'s boundary. Each block is a subcube, so the leaf MLE splits into a public selector times the block's own MLE, the padding contributing the leftover mass:
\[
  \widetilde V_0(\zeta)\;=\;\sum_b \eq(\mathsf{sel}_b,\zeta^{b}_{\mathrm{hi}})\Bigl(\gamma-\sum_{i=0}^{m-1}\alpha^{i}\,\widetilde c_{b,i}(\zeta^{b}_{\mathrm{lo}})\Bigr)\;+\;\Bigl(1-\sum_b\eq(\mathsf{sel}_b,\zeta^{b}_{\mathrm{hi}})\Bigr).
\]
The verifier forms the selectors $\eq(\mathsf{sel}_b,\cdot)$ and the constant coordinates itself. Each remaining $\widetilde c_{b,i}(\zeta^{b}_{\mathrm{lo}})$ is settled by its kind: a committed column by a direct opening (\S\ref{sec:prelim}), a virtual column off the committed columns it images (so $\gen\cdot\pc$ adds nothing beyond the $\pc$ opening), a \texttt{BLAKE3} value coordinate by a point evaluation of $q_\pkd$ at its slot (\S\ref{sec:tab-blake3}), the index column by \S\ref{sec:idxcol}. A committed column feeding several equal-size blocks is opened once, at their shared point $\zeta_{\mathrm{lo}}$.

\subsection{State interaction}\label{sec:state}

The state interaction carries $\tup{\dsep{ST},\pc,\fp}$. Each instruction row pulls its current state and pushes its successor $\tup{\dsep{ST},\mathit{next\_pc},\mathit{next\_fp}}$; the bus is preseeded by pushing the initial state $\tup{\dsep{ST},\pc_0,\fp_0}$ and pulling the final state $\tup{\dsep{ST},\pc_{\mathrm{final}},\fp_{\mathrm{final}}}$. When the bus balances, the per-row transitions form a chain from $(\pc_0,\fp_0)$ to $(\pc_{\mathrm{final}},\fp_{\mathrm{final}})$, possibly alongside disjoint cycles. The chain is the execution; a stray cycle is a closed loop of rows that still satisfies every constraint and reads the same committed memory and bytecode, so it cannot affect the start-to-final claim. By convention execution halts at a fixed sentinel: the last bytecode slot. Writing $N$ for the program length (a power of two, \S\ref{sec:e2e-bc}), the instructions sit at $\gen^{0},\dots,\gen^{\,N-1}$, so $\pc_{\mathrm{final}}=\gen^{\,N-1}$, with $\fp_{\mathrm{final}}=\gen^{0}$. A compiled \texttt{main} ends by jumping there. Both boundary states are then public, so the verifier forms them itself.

\section{Memory and bytecode lookups}\label{sec:omc}

Both a memory read and a bytecode read are indexed lookups into an array: the committed memory, or the public program. We realize each as an interaction on the shared bus (\S\ref{sec:m3}), balanced by \eqref{eq:gp}, using offline memory checking with a read count per index.

\subsection{Memory interaction}\label{sec:memchan}

The interaction carries $\tup{\dsep{MEM},\addr,\cnt,\val}$: the address and count are $\gen$-powers ($\K$-elements), the value a $128$-bit word. The word rides \emph{two} tuple coordinates, its two $\K$-lanes $(\val_0,\val_1)$, matching the two-lane committed memory image $M_0,M_1$; both are seeded, both pulled, so balance forces the whole word to match (below I write $\val$ for the pair). The tuple width is unchanged: the widest tuple is still the bytecode one (\S\ref{sec:bytecode}). The prover commits the memory $\mem$ ($2^{h}$ cells, two $\K$-lane columns) and the counts ("in the exponent", i.e. powers of $\gen$). With $A[i]$ the number of reads of address $i$, the flushes are:
\begin{itemize}
\item \textbf{Initialize:} for each address $i$, \push\ $\tup{\dsep{MEM},\gen^{i},1,\mem[i]}$ (seed count $\gen^{0}=1$).
\item \textbf{Read} (value $v$ at address $a$, count $\cnt=\gen^{t}$): \pull\ $\tup{\dsep{MEM},a,\cnt,v}$ and \push\ $\tup{\dsep{MEM},a,\gen\cdot\cnt,v}$.
\item \textbf{Finalize:} for each address $i$, \pull\ $\tup{\dsep{MEM},\gen^{i},\gen^{A[i]},\mem[i]}$, the committed final count $\gen^{A[i]}$.
\end{itemize}
\paragraph{Counts must be nonzero.} A read multiplies its count by $\gen$, pulling $\tup{\dsep{MEM},a,\cnt,v}$ and pushing $\tup{\dsep{MEM},a,\gen\cdot\cnt,v}$. If $\cnt=0$ these coincide (as $\gen\cdot0=0$) and cancel, so the read disappears from the bus and its $v$ is never checked against $\mem[a]$. Every count must therefore be nonzero, i.e.\ a power of $\gen$. A third grand product, over all count columns (memory and bytecode), enforces this: its root is nonzero iff every count is (\S\ref{sec:gkr}).

\paragraph{Counts must not wrap.} A count is the power $\gen^{t}$ after $t$ reads, so the total number of reads must stay below $\operatorname{ord}(\gen)=2^{64}-1$. At $64$ bits this is not waved through as automatic. The protocol fixes public \emph{instance caps}, and the verifier checks the announced instance against them before running any reduction (\S\ref{sec:e2e-unrolled}): memory size $16\le h\le 32$ (\S\ref{sec:vm}), each table's row count below $2^{32}$, and bytecode length at most $2^{32}$. A row makes at most $7$ reads (the \texttt{BLAKE3} row: six memory, one bytecode), so every \emph{admissible} instance performs
\[
  R\;\le\;7\cdot 6\cdot 2^{32}\;<\;2^{39}
\]
read flushes in total, far below $2^{64}-1$: the powers $\gen^{0},\gen^{1},\dots$ occurring in one address's history are pairwise distinct, and the counting arguments of this section are exact, with no probabilistic term. The caps cost no completeness ($2^{32}$ rows per table is far beyond any provable trace).

\paragraph{What balance certifies.} Fix an address $i$. Its only non-read flushes are the seed, a push of count $\gen^{0}$ with value $\mem[i]$, and the finalize, a pull of the committed final count with value $\mem[i]$.

\emph{Completeness.} An honest prover gives the $A[i]$ reads of $i$ the counts $\gen^{0},\dots,\gen^{A[i]-1}$ and the finalize the count $\gen^{A[i]}$. The read at count $\gen^{t}$ pulls $\gen^{t}$ and pushes $\gen^{t+1}$, so both sides are the multiset $\{\gen^{0},\dots,\gen^{A[i]}\}$ at value $\mem[i]$, and the bus balances.

\emph{Soundness.} Conversely, suppose the bus balances; we show every read of $i$ returns $\mem[i]$. Take any value $v\neq\mem[i]$. The seed and finalize carry $\mem[i]$, so the value-$v$ tuples come only from reads returning $v$, each pulling its count $\cnt$ and pushing $\gen\cdot\cnt$. Restricted to value $v$, balance equates the pulled with the pushed counts:
\[
  \{\,\cnt : \text{reads returning } v\,\}\;=\;\{\,\gen\cdot\cnt : \text{reads returning } v\,\}.
\]
This multiset of counts is thus invariant under multiplication by $\gen$. But the counts are powers of $\gen$, and $\times\gen$ cycles through all $2^{64}-1$ of them, so an invariant multiset weights every power equally: its size is a multiple of $2^{64}-1$. Under the instance caps its size is at most $R<2^{39}$, hence it is empty. So no read returns $v$, and every read of $i$ returns $\mem[i]$.

\subsection{Bytecode interaction}\label{sec:bytecode}

Instruction fetch is the same as the memory construction (\S\ref{sec:memchan}) with the \emph{public} program for the committed memory and the program counter for the address. The tuple is $\tup{\dsep{BC},\pc,\cnt,\mathit{instr}}$, where $\mathit{instr}$ is the encoded instruction: an opcode, three operand slots, and the two \texttt{DEREF} flags $(f_{\pc},f_{\fp})$, both $0$ for every opcode but \texttt{DEREF}.

\subsection{The index column}\label{sec:idxcol}

The memory seed and finalize tuples (\S\ref{sec:memchan}) run over every address $\gen^{i}$, $i=0,\dots,2^{h}-1$. Settling the grand product leaves the verifier evaluating, at a random $\zeta\in\E^{h}$, the MLE of this \emph{index column} $\gen^{0},\gen^{1},\dots,\gen^{2^{h}-1}$. It is not committed: as $\gen$-powers the MLE factors, costing $O(h)$.

Place address $\gen^{i}$ at the cube point $w\in\cube h$ of $i$'s bits, so $i=\sum_k 2^{k}w_k$ and
\[
  \gen^{\,i}\;=\;\prod_{k=0}^{h-1}\bigl(\gen^{\,2^{k}}\bigr)^{w_k}.
\]
Its MLE is the same product with each bit $w_k$ relaxed to $\zeta_k$,
\[
  \widetilde{\mathrm{idx}}(\zeta)\;=\;\prod_{k=0}^{h-1}\bigl(1+\zeta_k\,(1+\gen^{\,2^{k}})\bigr),
\]
each factor $1$ at $\zeta_k=0$ and $\gen^{2^{k}}$ at $\zeta_k=1$. The verifier evaluates it in $O(h)$ from the squares $\gen^{2^{0}},\gen^{2^{1}},\dots$. (Bytecode is the same, with the program counter for the address.)

\section{The instruction tables}\label{sec:tables}

Each instruction has its own table.

\subsection{\texttt{XOR}}\label{sec:tab-xor}

Asserts $\loc{o_C}=\loc{o_A}+\loc{o_B}$, the $128$-bit field sum.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$; operands $o_A,o_B,o_C$; addresses $a_A,a_B,a_C$; values $v_A,v_B,v_C$ (each two $\K$-lanes); memory counts $r_A,r_B,r_C$; bytecode count $r_{\mathrm{bc}}$.
\item \textbf{Constraints:} the three addresses $a_X=\fp\cdot o_X$ ($\K$), and the sum $v_C=v_A+v_B$ (the $\E$-relation on the value lanes, reassembled $v=v_0+v_1 y$; lane-wise it is two $\K$-equations).
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\gen\cdot\pc,\fp}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{XOR},o_A,o_B,o_C,0,0}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$.
  \item \textbf{Memory} ($X\in\{A,B,C\}$): \pull\ $\tup{\dsep{MEM},a_X,r_X,v_X}$, \push\ $\tup{\dsep{MEM},a_X,\gen\cdot r_X,v_X}$.
  \end{itemize}
\end{itemize}

\subsection{\texttt{MUL\_NATIVE}}\label{sec:tab-mul}

Identical to \texttt{XOR} but for the product:
\begin{itemize}
\item \textbf{Columns:} as \texttt{XOR}.
\item \textbf{Constraints:} the three addresses and $v_C=v_Av_B$ in $\E$. With $y^2=x y+1$,
\[
v_{C,0}=v_{A,0}v_{B,0}+v_{A,1}v_{B,1},\qquad
v_{C,1}=v_{A,0}v_{B,1}+v_{A,1}v_{B,0}+x\,v_{A,1}v_{B,1}.
\]
\item \textbf{Flushes:} as \texttt{XOR}, with opcode $\opc{MUL}$.
\end{itemize}

\subsection{\texttt{SET\_CONSTANT}}\label{sec:tab-set}

Asserts $\loc{o}=k$, how a constant enters the read-only memory: the row ties the committed cell to the bytecode immediate $k$.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$; operand $o$; immediate $k$ ($128$-bit, two $\K$-lanes $k_0,k_1$); address $a$; memory count $r$; bytecode count $r_{\mathrm{bc}}$.
\item \textbf{Constraints:} the address $a=\fp\cdot o$.
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\gen\cdot\pc,\fp}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{SET},o,k_0,k_1,0,0}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$ (the immediate's two lanes ride the spare operand slots $5,6$).
  \item \textbf{Memory:} \pull\ $\tup{\dsep{MEM},a,r,k}$, \push\ $\tup{\dsep{MEM},a,\gen\cdot r,k}$ (the value slot is the constant $k$, its two lanes).
  \end{itemize}
\end{itemize}

\subsection{\texttt{DEREF}}\label{sec:tab-deref}

\texttt{DEREF} follows a pointer and asserts the cell it points to. The row reads three cells:
\begin{itemize}
\item the \emph{pointer} $p$, at $a_1=\fp\cdot\alpha$;
\item the \emph{target} $v_2$, at the dereferenced address $a_2=p\cdot\beta$;
\item the \emph{local} value $v_3$, at $a_3=\fp\cdot\gamma$.
\end{itemize}
The store asserts that $v_2$ equals the source chosen by the mode $s$ (\S\ref{sec:vm}): the local cell $v_3$, the return address $\gen^{2}\cdot\pc$ (a virtual $\times\gen^{2}$ of $\pc$, \S\ref{sec:prelim}), or the frame pointer $\fp$. The mode is two boolean flags $(f_{\pc},f_{\fp})$: $(0,0)$ for $\texttt{deref\_cell}[\gamma]$, $(1,0)$ for $\texttt{deref\_pc}$, $(0,1)$ for $\texttt{deref\_fp}$.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$; operands $\alpha,\beta,\gamma$; flags $f_{\pc},f_{\fp}$; addresses $a_1,a_2,a_3$; pointer $p$ (a memory word, two $\K$-lanes); values $v_2,v_3$ (each two $\K$-lanes); memory counts $r_1,r_2,r_3$; bytecode count $r_{\mathrm{bc}}$.
\item \textbf{Constraints:} the addresses $a_1=\fp\cdot\alpha$, $a_2=p\cdot\beta$, $a_3=\fp\cdot\gamma$, and the store
\[
  v_2 \;=\; (1+f_{\pc}+f_{\fp})\,v_3 \;+\; f_{\pc}\,(\gen^{2}\cdot\pc) \;+\; f_{\fp}\,\fp ,
\]
which gives $v_3$, $\gen^{2}\cdot\pc$, or $\fp$ at the three flag settings (in characteristic $2$ the $v_3$ term zeroes for the last two). The store is an $\E$-relation on the value lanes; $a_2=p\cdot\beta$, with $a_2$ a single $\K$ column and $\beta\in\K$, forces the pointer word $p$ into $\K$ (a genuine $\gen$-power). The flags need no validity check: the verifier reads them from the public program (\S\ref{sec:e2e-bc}).
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\gen\cdot\pc,\fp}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{DRF},\alpha,\beta,\gamma,f_{\pc},f_{\fp}}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$.
  \item \textbf{Memory:} three reads, each \pull\ at count $r$ and \push\ at $\gen\cdot r$: pointer $\tup{\dsep{MEM},a_1,r_1,p}$, target $\tup{\dsep{MEM},a_2,r_2,v_2}$, local cell $\tup{\dsep{MEM},a_3,r_3,v_3}$.
  \end{itemize}
\end{itemize}

\subsection{\texttt{JUMP}}\label{sec:tab-jump}

A conditional jump on whether the condition $c=\loc{o_c}$ is nonzero, with $c$ an arbitrary $128$-bit word. It transfers to $\pc\gets d$, $\fp\gets f$ when $c\neq0$ (the destination $d=\loc{o_d}$, a code address, and the new frame $f=\loc{o_f}$), and falls through, $\pc\gets\gen\cdot\pc$, $\fp\gets\fp$, when $c=0$. The branch is taken according to a committed boolean indicator $b=[\,c\neq0\,]$, which a prover-supplied inverse $w$ certifies through the two degree-$2$ constraints
\[
  b=c\cdot w,\qquad c\,(b+1)=0 .
\]
When $c\neq0$ the second forces $b=1$ and the first then $w=c^{-1}$; when $c=0$ the first forces $b=0$. Writing $s=\gen\cdot\pc$ for the virtual fall-through, the committed next state is the selection
\[
  \mathit{next\_pc}=b\cdot d+(b+1)\,s,\qquad \mathit{next\_fp}=b\cdot f+(b+1)\,\fp.
\]
Alone among the tables, \texttt{JUMP} commits its successors.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$ and $\mathit{next\_pc},\mathit{next\_fp}$ (single $\K$-columns, the next state); operands $o_c,o_d,o_f$; addresses $a_c,a_d,a_f$; the condition $c$ and the destination/frame words $d,f$ (each two $\K$-lanes); inverse $w$ (two $\K$-lanes, $c^{-1}$ in $\E$); indicator $b$ (one $\K$-lane, $0/1$); memory counts $r_c,r_d,r_f$; bytecode count $r_{\mathrm{bc}}$. ($w$ and $b$ are witness columns, neither read from memory nor in the bytecode; the fall-through $s=\gen\cdot\pc$ is virtual.)
\item \textbf{Constraints:} the three addresses $a_X=\fp\cdot o_X$; the indicator pair $b=c\cdot w$ and $c\,(b+1)=0$ ($\E$-relations on the lanes); and the two selections $\mathit{next\_pc},\mathit{next\_fp}$ above. Since $\mathit{next\_pc},\mathit{next\_fp}$ are single $\K$-columns, the selections force the chosen words $d,f$ into $\K$ on a taken branch.
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\mathit{next\_pc},\mathit{next\_fp}}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{JMP},o_c,o_d,o_f,0,0}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$.
  \item \textbf{Memory} ($X\in\{c,d,f\}$): \pull\ $\tup{\dsep{MEM},a_X,r_X,X}$, \push\ $\tup{\dsep{MEM},a_X,\gen\cdot r_X,X}$.
  \end{itemize}
\end{itemize}

\subsection{\texttt{PACK64X2}}\label{sec:tab-pack64x2}

Asserts that two source cells are $\K$-valued and packs them into one canonical $128$-bit word.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$; operands $o_A,o_B,o_C$; addresses $a_A,a_B,a_C$; source limbs $v_A,v_B\in\K$; memory counts $r_A,r_B,r_C$; bytecode count $r_{\mathrm{bc}}$.
\item \textbf{Constraints:} the three addresses $a_X=\fp\cdot o_X$. Value validity is enforced by the memory interactions rather than separate AIR equations.
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\gen\cdot\pc,\fp}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{PACK},o_A,o_B,o_C,0,0}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$.
  \item \textbf{Memory:} read $\tup{\dsep{MEM},a_A,r_A,v_A,0}$ and $\tup{\dsep{MEM},a_B,r_B,v_B,0}$, then read the destination as $\tup{\dsep{MEM},a_C,r_C,v_A,v_B}$; each pull is paired with the corresponding push at count $\gen\cdot r$. The literal zeros make the two source range assertions and the destination packing exact through bus balance.
  \end{itemize}
\end{itemize}

\subsection{\texttt{BLAKE3}}\label{sec:tab-blake3}

Compresses $64$ input bytes to the $32$ output bytes at $c$. Each $128$-bit chunk occupies one cell. The four input addresses are $a_i=\fp\cdot o_i$ for $i\in\{0,1,2,3\}$ and need not be adjacent. The output occupies $a_C=\fp\cdot o_C$ and $\gen\cdot a_C$. Thus each row accesses four input cells and two output cells.

Flock~\cite{flock} proves $c=\mathrm{BLAKE3}(a,b)$ with a batched R1CS over a Boolean witness. The witness is packed into $q_\pkd$, with $64$ bits per $\K$-element. Ring switching~\cite{cryptoeprint:2024/504} maps its $\F_2$ claims to one $\E$-weighted claim on $q_\pkd$, which joins the batched opening (\S\ref{sec:e2e-unrolled}).

The words $a,b,c$ are contained in $q_\pkd$. Each $64$-bit flock word occupies one packed slot, and each $128$-bit cell occupies two slots. The twelve lanes $v_{A0},\dots,v_{C3}$ therefore need no separate commitment. Each lane is a strided point evaluation of $q_\pkd$. The R1CS matrices fix the BLAKE3 IV, counter, block length, and flags, so they prove the compression of $a\Vert b$ without separate pin openings.
\begin{itemize}
\item \textbf{Columns:} $\pc,\fp$; operands $o_0,o_1,o_2,o_3,o_C$; input/output addresses $a_0,a_1,a_2,a_3,a_C$; per-cell memory counts $r_0,\dots,r_3$ (the four input cells) and $r_{C0},r_{C1}$ (the two output cells); bytecode count $r_{\mathrm{bc}}$. The twelve value lanes $v_{A0},\dots,v_{C3}$ (two per value cell) are committed "inside" $q_\pkd$, see above. The successor address $\gen\cdot a_C$ is virtual.
\item \textbf{Constraints:} the five degree-$2$ relations $a_i=\fp\cdot o_i$ for $i\in\{0,1,2,3\}$ and $a_C=\fp\cdot o_C$. Flock proves $c=\mathrm{BLAKE3}(a,b)$.
\item \textbf{Flushes:}
  \begin{itemize}
  \item \textbf{State:} \pull\ $\tup{\dsep{ST},\pc,\fp}$, \push\ $\tup{\dsep{ST},\gen\cdot\pc,\fp}$.
  \item \textbf{Bytecode:} \pull\ $\tup{\dsep{BC},\pc,r_{\mathrm{bc}},\opc{B3},o_0,o_1,o_2,o_3,o_C}$, \push\ the same at $\gen\cdot r_{\mathrm{bc}}$.
  \item \textbf{Memory:} six reads, each \pull\ at count $r$ and \push\ at $\gen\cdot r$, its value the cell's two $\K$-lanes (\S\ref{sec:memchan}). The four input cells \pull\ $\tup{\dsep{MEM},a_i,r_i,v_i}$; the output \pull s $\tup{\dsep{MEM},a_C,r_{C0},v_{C0}}$ and $\tup{\dsep{MEM},\gen\cdot a_C,r_{C1},v_{C1}}$.
  \end{itemize}
\end{itemize}

\section{End-to-end protocol}\label{sec:e2e}

This section assembles the primitives into the executable protocol, beginning with the constants, encodings, and padding fixed in advance.

\subsection{Constants}\label{sec:e2e-const}

Fixed once, public, and shared by prover and verifier: the fields, the bus shape, and the tags naming interactions and instructions on it.

\paragraph{Fields, generator, and bus.}
\[
\K=\F_2[x]/(x^{64}+x^4+x^3+x+1),\qquad
\E=\K[y]/(y^2+x y+1),\qquad
\gen=x,\qquad \operatorname{ord}(\gen)=2^{64}-1.
\]
Committed columns are $\K$-valued; machine words and challenges are $\E$-valued (\S\ref{sec:prelim}). Every bus tuple has width $m=9$ (the bytecode tuple); shorter tuples are zero-padded in their high coordinates (\S\ref{sec:m3}).

\paragraph{Domain separators.} Coordinate $0$ of every tuple is a domain separator, naming the interaction it belongs to (\S\ref{sec:m3}):
\[
  \dsep{ST}=1,\qquad \dsep{MEM}=\gen,\qquad \dsep{BC}=\gen^{2}.
\]

\paragraph{Opcodes.} Coordinate $3$ of every bytecode tuple is an opcode, naming the instruction stored at that program counter (\S\ref{sec:bytecode}):
\[
  \opc{XOR}=1,\quad \opc{MUL}=\gen,\quad \opc{SET}=\gen^{2},\quad \opc{DRF}=\gen^{3},\quad \opc{JMP}=\gen^{4},\quad \opc{B3}=\gen^{5},\quad \opc{PACK}=\gen^{6}.
\]

\subsection{Bytecode encoding}\label{sec:e2e-bc}

The program is public. Instruction $\pc$ fills coordinates $3$ through $8$ of its bytecode tuple $\tup{\dsep{BC},\pc,\cnt,\cdot}$. Coordinate $3$ holds the opcode, and coordinates $4,5,6$ hold three operands. Coordinates $7,8$ hold the \texttt{DEREF} flags $(f_{\pc},f_{\fp})$, the \texttt{BLAKE3} operands $(o_3,o_C)$, or zero. Each operand is a $\gen$-power or an immediate. For \texttt{SET\_CONSTANT}, coordinates $5,6$ hold the two $\K$-lanes of the immediate.
\begin{center}
\begin{tabular}{lllllll}
\hline
Instruction & coord.\ $3$ & coord.\ $4$ & coord.\ $5$ & coord.\ $6$ & coord.\ $7$ & coord.\ $8$\\
\hline
\texttt{XOR}           & $\opc{XOR}$ & $o_A$ & $o_B$ & $o_C$ & $0$ & $0$\\
\texttt{MUL\_NATIVE}   & $\opc{MUL}$ & $o_A$ & $o_B$ & $o_C$ & $0$ & $0$\\
\texttt{SET\_CONSTANT} & $\opc{SET}$ & $o$   & $k_0$ & $k_1$ & $0$ & $0$\\
\texttt{DEREF}         & $\opc{DRF}$ & $\alpha$ & $\beta$ & $\gamma$ & $f_{\pc}$ & $f_{\fp}$\\
\texttt{JUMP}          & $\opc{JMP}$ & $o_c$ & $o_d$ & $o_f$ & $0$ & $0$\\
\texttt{PACK64X2}      & $\opc{PACK}$ & $o_A$ & $o_B$ & $o_C$ & $0$ & $0$\\
\texttt{BLAKE3}        & $\opc{B3}$  & $o_0$ & $o_1$ & $o_2$ & $o_3$ & $o_C$\\
\hline
\end{tabular}
\end{center}
So the program is six public columns (opcode, three operands, two flags), whose MLEs the verifier forms itself. Writing $\mathit{instr}$ for the six, the bytecode seed at $\pc$ pushes $\tup{\dsep{BC},\pc,1,\mathit{instr}}$ and an instruction table pulls it with its opcode hardcoded (\S\ref{sec:tables}); balance then forces each table's operands \emph{and} flags to match the program at its $\pc$.

The program is padded to a power of two (\S\ref{sec:idxcol}). A padding entry is never executed: no table pulls it, so its seed push and finalize pull (both at count $1$) cancel.

\subsection{Padding rows}\label{sec:e2e-pad}

Each table is padded to a power of two. Padding rows are a fixed default, not all zeros (a zero count would self-cancel a read, \S\ref{sec:memchan}): every column zero but the read counts $\gen^{0}=1$, satisfying every constraint. The prover announces each table's real row count and the verifier divides the default surplus out of the bus product (\S\ref{sec:gp}).

\subsection{Public input}\label{sec:e2e-pi}

The public input is the first two memory cells $\mem[0],\mem[1]$, each a $128$-bit word, fixed and agreed by both parties. Cells $0$ and $1$ span the subcube of the address cube $\cube h$ with the $h-1$ high coordinates $0$ and the low one free, so on \emph{each} committed memory lane $M_\ell$ ($\ell\in\{0,1\}$, \S\ref{sec:memchan}) the MLE restricts there to the line $\widetilde{\mathrm{PI}}_\ell(r)=(1-r)\,\mem[0]_\ell+r\,\mem[1]_\ell$, which the verifier forms itself from the public words. It samples $r_m\in\E$ and checks both lanes,
\[
  \widetilde M_\ell(r_m,0,\dots,0)\;=\;\widetilde{\mathrm{PI}}_\ell(r_m),\qquad \ell\in\{0,1\}.
\]
Writing $P=P_0+P_1y$ for the public-input evaluation, the implementation transmits only $P_0$ and computes
\[
P_1=(P+P_0)y^{-1}.
\]
The final stacked opening discharges both lane claims.

\subsection{The unrolled protocol}\label{sec:e2e-unrolled}

We lay out the whole protocol as one ordered sequence. Each reduction below adds evaluation claims on committed columns to a shared \emph{claim pool}; the final \textsc{Opening} phase proves all of them with a single PCS opening.

\paragraph{Setup and statement binding.} Fixed and shared in advance (\S\ref{sec:e2e-const} through \S\ref{sec:e2e-pi}): the fields and constants; the instance caps (\S\ref{sec:memchan}); the public program and its length; the initial state $(\pc_0,\fp_0)$ and final state $(\pc_{\mathrm{final}},\fp_{\mathrm{final}})$; and the public input $\mem[0],\mem[1]$. Before any challenge, the transcript is seeded by the public input and one environment digest that binds both the exact bytecode and flock's BLAKE3 circuit family. The prover then announces $h$ and the seven real opcode-row counts; the verifier checks them against the caps and derives every table, stack, and leaf shape. Thus neither a different program nor an adaptively chosen shape can reuse the proof's later challenges.

\paragraph{Commitment.}
\begin{enumerate}
\item \textbf{Prover.} Builds all committed columns ($\K$-valued; a $128$-bit value is two $\K$-lane columns, \S\ref{sec:prelim}): each table's state, operands, addresses, read-value lanes, and counts (plus $w,b,\mathit{next\_pc},\mathit{next\_fp}$ for \texttt{JUMP}); the two memory lanes $M_0,M_1$; the memory and bytecode finalize counts; and flock's packed witness $q_\pkd$ (\S\ref{sec:tab-blake3}), always present (a program that executes no \texttt{BLAKE3} carries one padding instance, keeping the proof shape uniform). Counts are always committed $\gen$-powers (\S\ref{sec:memchan}); the public program and the virtual increments $\gen\cdot\pc,\gen\cdot\cnt$ are not. It stacks these into $q$, sends the dense commitment (\S\ref{sec:stacking}, \S\ref{sec:rs-dense}), and announces each table's non-padded row count, which the verifier checks against the instance caps (\S\ref{sec:memchan}) and rounds up to a power of two to fix the stack and leaf layouts (\S\ref{sec:leafstack}). Padding rows are a fixed default (\S\ref{sec:e2e-pad}); the announced sizes fix the surplus the verifier later divides out (\S\ref{sec:gp}).
\end{enumerate}

\paragraph{Bus.}
\begin{enumerate}
\item \textbf{Verifier.} Sends the fingerprint challenge $\alpha$ and the multiset challenge $\gamma$.
\item \textbf{Prover.} Forms the push, pull, and \emph{count} leaf vectors (\S\ref{sec:leafstack}; the count leaves are the count columns themselves, \S\ref{sec:memchan}), stacked into blocks and padded with $1$s. Sends the count root $R_c$ and one bus root $R$, the push/pull product after the default surplus is removed (\S\ref{sec:gp}).
\item \textbf{Prover \& Verifier.} Run the one batched GKR (\S\ref{sec:gkr}) over all three trees. For the bus, the verifier multiplies $R$ by each side's public surplus and checks the side's product against it; one $R$ for both sides is the balance check (\S\ref{sec:gp}). For the counts it checks $R_c\neq0$ (no read self-cancels, \S\ref{sec:memchan}). The pass yields three leaf claims $\widetilde V^{t}_0(\zeta)$ at one shared $\zeta$.
\item \textbf{Prover \& Verifier.} Decompose each leaf claim (\S\ref{sec:leafstack}). The verifier forms the public part: the block selectors; the constant coordinates (separators, the per-table read opcodes, the boundary state $(\pc_0,\fp_0)$ and $(\pc_{\mathrm{final}},\fp_{\mathrm{final}})$); the index column (\S\ref{sec:idxcol}); and the public program columns of the bytecode seed and finalize, whose MLEs it forms directly. The prover supplies each block's committed-column evaluations (one $\E$-element each, \S\ref{sec:prelim}); a virtual coordinate such as $\gen\cdot\pc$ is read off its committed source, and a \texttt{BLAKE3} value coordinate off $q_\pkd$ at its slot (\S\ref{sec:tab-blake3}). The verifier checks the decomposition formula against $\widetilde V_0(\zeta)$; the supplied evaluations enter the claim pool. The count channel decomposes likewise (its leaves are the count columns), adding only their evaluations.
\end{enumerate}

\paragraph{Local constraints.} Run once per table (\S\ref{sec:air}); the seven runs share no challenges.
\begin{enumerate}
\item \textbf{Verifier.} For the table, sends the constraint-batch challenge $\eta\in\E$ and the zerocheck point $r\in\E^{\tau}$, where $2^{\tau}$ is its row count.
\item \textbf{Prover \& Verifier.} Run the table's zerocheck, a $\tau$-round sumcheck on the $\eta$-batched constraint $\sum_x C(x)\,\eq(r,x)=0$. Its end needs that table's (committed) column evaluations at the random point, which enter the pool.
\end{enumerate}

\paragraph{Public input.}
\begin{enumerate}
\item \textbf{Verifier.} Sends $r_m$, forms $\widetilde{\mathrm{PI}}_\ell(r_m)$ from the public input, and adds to the pool the two memory-lane claims $\widetilde M_\ell(r_m,0,\dots,0)=\widetilde{\mathrm{PI}}_\ell(r_m)$, $\ell\in\{0,1\}$ (\S\ref{sec:e2e-pi}).
\end{enumerate}

\paragraph{BLAKE3 validity.}
\begin{enumerate}
\item \textbf{Prover \& Verifier.} Reduce the executed BLAKE3 constraints to two evaluation claims on $q_\pkd$ (\S\ref{sec:tab-blake3}).
\item \textbf{Prover \& Verifier.} The ring switching (\S\ref{sec:tab-blake3}) reduces each bit-witness claim to a weighted-sum claim on the $q_\pkd$ region of the stack, with a transparent $\E$-valued weight the verifier can evaluate; both join the pool.
\end{enumerate}

\paragraph{Opening.}
\begin{enumerate}
\item \textbf{Prover \& Verifier.} Each pooled claim is one evaluation of a committed column at a point, its value already supplied; via the stacking selectors (\S\ref{sec:stacking}) it becomes a weighted sum $\sum_w W_j(w)\,\widehat q(w)=c_j$ over the stacked witness. (The ring-switched claims arrive already in this form, their weights supported on the $q_\pkd$ region.)
\item \textbf{Verifier.} Sends the batching coefficients $\lambda_1,\dots,\lambda_J\in\E$, one per claim.
\item \textbf{Prover \& Verifier.} Fold the $J$ claims into $W_\lambda=\sum_j\lambda_j W_j$ and $C_\lambda=\sum_j\lambda_j c_j$, and run the inner-product PCS opening on $\sum_w W_\lambda(w)\,\widehat q(w)=C_\lambda$ (\S\ref{sec:rs-dense}), the verifier evaluating $W_\lambda$ itself. This one opening discharges every pooled claim; there is no separate reduction sumcheck.
\item \textbf{Verifier.} Accepts iff the announced sizes passed the caps, the two roots agreed, each leaf decomposition and every sumcheck's final value matched the supplied evaluations, flock's reduction verified, the public-input claim held, and the PCS opening verified.
\end{enumerate}

\section{Recursive aggregation}\label{sec:recursion}

The implemented N-to-1 aggregation runs the verifier of \S\ref{sec:e2e-unrolled} \emph{inside} the VM: one compiled guest replays the transcripts, sumchecks, ring switches, and stacked Ligerito-K openings of $n_{\mathrm{rec}}\ge1$ ordinary VM proofs for a fixed inner program. The proofs may have different announced table and commitment sizes; the guest dispatches among the certified shapes without recompilation. The VM is well suited to this because $\E$ is its machine-word field and its Fiat--Shamir hash is the \texttt{BLAKE3} opcode. One class of operations, however, does not fit.

\subsection{Deferred evaluation claims}\label{sec:deferred-claims}

Twice during verification, a large \emph{fixed} multilinear must be evaluated at a random point:
\begin{itemize}
\item \textbf{the bytecode}: during leaf decomposition (\S\ref{sec:e2e-unrolled}, \emph{Bus}), the verifier evaluates the stacked polynomial formed from the six public program columns and three selector variables, a pass over the whole program;
\item \textbf{the flock matrices}: the \texttt{BLAKE3} relation is proven by flock (\S\ref{sec:tab-blake3}), whose lincheck ends by evaluating $\widetilde A_0$ and $\widetilde B_0$, the per-block R1CS matrices: $28$-variable multilinears with $\approx 21\mathrm{M}$ nonzero Boolean entries between them.
\end{itemize}
Either evaluation is one pass over a fixed object. This takes milliseconds natively, but in-circuit it would dwarf the entire remaining verifier, in cycles and in bytecode size alike.

The aggregation guest therefore does not evaluate these fixed tables. It verifies every surrounding reduction and exports only the resulting points and claimed values.

There are three fixed polynomials: the stacked bytecode and the two matrices $A_0,B_0$. Verifying each ordinary sub-proof produces one claim on each, for $3n_{\mathrm{rec}}$ claims. A fresh Fiat--Shamir transcript absorbs all sub-statements and claim data, then batches them with two sumchecks: one for bytecode and one for the matrices, which share their $28$ variables. The guest reduces the batch to one bytecode evaluation and a common-point evaluation of each matrix. These three reduced claims, the inner proving-environment digest, and the sub-statements are hashed into the outer VM's two-word public input.

The public \texttt{RecursiveProof::verify} path first verifies that outer VM proof and then evaluates the three fixed polynomials natively at the reduced points. It rejects an empty batch and is the only complete acceptance path. Feeding a \texttt{RecursiveProof} into another aggregation layer (carrying its reduced claims forward rather than checking them natively) is not yet exposed by the current harness.

The matrix sumcheck runs over the $28$-variable polynomial
\[
  P(x) \;=\; A_0(x)\,W_A(x) \;+\; B_0(x)\,W_B(x),
  \qquad
  W_M(x) \;=\; \textstyle\sum_t \gamma_{M,t}\,\eq(r_t,x),
\]
and ends in the reduced claims $\widetilde A_0(r^\ast)$ and $\widetilde B_0(r^\ast)$, at a common point. The in-circuit cost is that of the sumcheck verifier: $28$ cheap rounds. The prover exploits sparsity to avoid the $2^{28}$-sized cube. During the first $14$ rounds (the row variables), each claim enters through the column contraction $m_t[i]=\sum_{j\in\mathrm{row}_i}\eq(r_t^{\mathrm{col}},j)$; for the last $14$ rounds, each matrix is contracted once along its rows, at the row point just fixed. Each contraction is a single addition-only pass over the nonzeros (the entries are Boolean), and all remaining work is dense of size $2^{14}$. Batching $t$ claims per matrix therefore costs $t+1$ passes over its nonzeros plus $O(t\cdot2^{14})$ multiplications, against the $O(2^{28})$ field operations and memory of a dense sumcheck. The bytecode sumcheck is the same, only smaller and with a dense polynomial.

\section{ISA programming}\label{sec:isa-programming}

How the zkDSL compiler (\texttt{crates/lean\_compiler}; surface language in \texttt{crates/lean\_compiler/zkDSL.md}) programs the seven-instruction ISA. Two themes run through every pattern: \emph{write-once memory is an assertion mechanism} (a second write of a cell is an equality check, an unwritten cell is prover-chosen), and \emph{indices live in the exponent} (a logical index $i$ is the element $\gen^{i}$, so incrementing, offsetting, and address formation are single field multiplications).

\subsection{Hints}\label{sec:prog-hints}

The prover's nondeterminism enters through write-once cells that are still unset when execution needs them. Some are explicit advice (a fresh allocation pointer $\gen^{\mathrm{base}}$, a witness stream, a computed-advice builtin); others are back-solved by an instruction whose remaining operands are known. In every case the resulting word is unconstrained except by the instructions and memory equalities that consume it, so a program must check all security-relevant advice.

\subsection{Division and inequality}\label{sec:prog-div-ne}

The zkDSL's single-slash division $q=a/b$ uses write-once back-solving: it emits \texttt{MUL\_NATIVE} with $q$ as the one unset input and $a$ as the already-written output. Witness generation fills $q=a\,b^{-1}$, while the ordinary multiplication table proves $q b=a$. It costs one VM instruction. Division by zero is undefined. Double slash \texttt{//} remains compile-time integer floor division for sizes and indices; it is not a field operation.

The assertion \texttt{assert a != b} first computes $a+b$ with \texttt{XOR}. A conditional \texttt{JUMP} skips a poison path exactly when this value is nonzero. Equality falls through to a jump to $\gen^{-1}$, outside the committed bytecode cube, so the bytecode bus cannot balance any purported continuation. This proves inequality without an inverse hint.

\subsection{Functions}\label{sec:prog-functions}

A frame is fp-relative with layout $[\mathit{retpc},\mathit{retfp},\mathrm{args},\mathrm{rets},\mathrm{locals}]$. A call: the callee frame pointer is hinted into a fresh cell $\mathit{nfp}$; the arguments and the caller's $\fp$ are stored through $\mathit{nfp}$ with \texttt{DEREF}s (modes \texttt{cell} and \texttt{fp}); a \texttt{DEREF} in mode \texttt{pc} stores the return address $\gen^{2}\cdot\pc$ (the instruction after the transfer); one \texttt{JUMP} enters the callee. Return is a single \texttt{JUMP} to $\loc{1}$ with frame pointer $\loc{\gen}$. Cost: $n_{\mathrm{args}}+n_{\mathrm{rets}}+4$ instructions per call.

\subsection{Loops}\label{sec:prog-loops}

A counted loop keeps its counter in the exponent: it starts at $\gen^{\mathrm{lo}}$, advances by one multiplication $i\gets\gen\cdot i$, and exits on reaching $\gen^{\mathrm{hi}}$. The body is lowered as a tail-recursive helper function whose exit test \emph{is} the recursive call's \texttt{JUMP} condition (the \texttt{XOR} $i+\gen^{\mathrm{hi}}$ is nonzero exactly while the loop must continue), so an iteration costs one call plus two instructions, with no is-zero gadget. Loop state is threaded through captured arguments or a heap buffer.

\subsection{Conditionals}\label{sec:prog-conditionals}

\texttt{if}/\texttt{else} on a field equality is one \texttt{XOR} plus one conditional \texttt{JUMP}: the taken jump goes to whichever block the test should \emph{not} fall into ($=$ falls into \emph{then}, $\neq$ into \emph{else}), so no negation is ever computed. Intra-function jump targets are bytecode addresses $\gen^{\mathrm{entry}+i}$, backpatched at layout.

A taken \texttt{JUMP} reloads $\fp$ from a cell, and the ISA has no $\fp$-read; a branching function therefore materializes its own $\fp$ once: a \texttt{DEREF} in mode \texttt{fp} writes $\fp$ into a fresh one-cell heap buffer, a \texttt{DEREF} in mode \texttt{cell} copies it back into the frame (2 instructions, amortized; in \texttt{main}, $\fp=\gen^{0}=1$ is already a constant). Bindings made inside a branch are local to it; branches communicate through write-once cells: only one branch executes, so both may write the \emph{same} cell.

\subsection{Range checks}\label{sec:prog-range-checks}

The range check \emph{in the exponent} proves $x\in\{\gen^{0},\dots,\gen^{k-1}\}$, i.e.\ $\log_{\gen}x<k$, in 3 cycles (leanVM's \texttt{DEREF} trick \cite{leanvm-doc}, transported to $\gen$-powers):
\begin{enumerate}
\item \texttt{DEREF} through $x$: the dereferenced address $x\cdot\gen^{0}$ rides the memory interaction, which balances only for the $2^{h}$ addresses $\gen^{0},\dots,\gen^{2^{h}-1}$ (\S\ref{sec:memchan}): the bus itself proves $x=\gen^{e}$ with $e<2^{h}$;
\item \texttt{MUL} $x\cdot y$ into a write-once cell holding the constant $\gen^{k-1}$: the runner back-solves the complement $y=\gen^{k-1-e}$ (\S\ref{sec:prog-hints}), and the double write asserts $x\cdot y=\gen^{k-1}$;
\item \texttt{DEREF} through $y$: proves $y=\gen^{f}$, $f<2^{h}$.
\end{enumerate}
Then $e+f\equiv k-1\pmod{2^{64}-1}$ with $e,f<2^{h}\le2^{32}$, so $e+f<2^{33}<2^{64}-1$ and the congruence is an equality over the integers; a ``negative'' $k-1-e$ would wrap to $\approx2^{64}\gg2^{h}$, so $e\le k-1$, for every memory size the prover may announce, provided $k\le2^{16}$ (the minimum, \S\ref{sec:vm}). The two \texttt{DEREF} targets are deferred touches; the constant cell is one amortized \texttt{SET} per distinct bound per frame.

\subsection{Match statements}\label{sec:prog-match}

leanVM dispatches a switch with a single jump to $\pc=a+b\cdot x$: the case blocks, padded to a common length $b$, sit consecutively from base $a$, and its integer arithmetic forms the affine address in one instruction. That computation does not transport to this VM. Addresses here are $\gen$-powers: the additive offset $a+j$ is the cheap product $\gen^{a}\cdot x$, but the \emph{scaling} $b\cdot j$ is the power $x^{b}$ ($\log_2 b$ squarings per dispatch), with every block still padded to the longest. Instead, the aligned region collapses to fixed two-instruction \emph{trampoline slots} and the real blocks move out of line: matching on $x=\gen^{j}$, $j\in[0,n)$, the dispatch jumps to
\[
  d \;=\; \gen^{T}\cdot x^{2} \;=\; \gen^{T+2j},
\]
slot $j$ of a table at bytecode base $T$, where a slot is $\texttt{SET}\ c=\gen^{\mathrm{block}_j}$ then $\texttt{JUMP}\ c$. The case blocks sit anywhere, unaligned and of any length; the scaling cost is the single squaring $x^{2}$. Two jumps per dispatch, $\approx7$ cycles independent of $n$.

The slots are two instructions rather than one because a \texttt{JUMP} reads its destination from a \emph{cell}: a one-instruction slot would need its target cell pre-written, i.e.\ $n$ \texttt{SET}s executed before every dispatch. (Other layouts exist and trade exactly this, e.g.\ a memory-resident table of the $n$ block addresses, read with one \texttt{DEREF} and taken with a \emph{single} jump, paying the \texttt{SET}s as setup; worthwhile for many repeated small matches. The compiler currently emits only the trampoline.)

As in leanVM, the matched value must be known to lie in $[0,n)$: a hinted $x$ must be range-checked first (\S\ref{sec:prog-range-checks}); otherwise the dispatch lands at an attacker-chosen program counter.

\iffalse % Deferred annex (u32 addition gadget, ring-switching): commented out for now.
% NOTE: this annex predates the 64-bit transition and is written for the old
% F_{2^128} design (D = 128, nu = 7); re-derive over K/E before reviving it.
\appendix

\section{The \texttt{u32} addition gadget (deferred)}\label{sec:add}

This appendix is \emph{unused by v1}; it is kept for the deferred integer instructions \texttt{u32\_ADD} and \texttt{IS\_NONZERO} (\S\ref{sec:vm}), which need genuine $32$-bit addition with carries. Over $\F_{2^{128}}$ that integer sum is not the field sum, so it is proven bit-level by the gadget below, run as a single \emph{addition service}: one table proving a batch of additions $a+b=c\bmod2^{32}$ and pulling each proven row as $\tup{\dsep{ADD},a,b,c,e}$ on a fourth domain $\dsep{ADD}$, with $e\in\F_2$ the overflow (carry out of the top bit); consumers push the tuple they need, and balance forces every pushed addition to equal a proven one. Reviving it restores the bit-columns of \S\ref{sec:prelim} (a word as its $32$ bit-columns, materialized $\sum_j x^j\widetilde b_j$) and ring-switching (Appendix~\ref{sec:ringswitch}) to commit them. \texttt{IS\_NONZERO} reads off $e$ from $x+(2^{32}{-}1)$, whose overflow is $[x\neq0]$.

The gadget proves $C=A+B$ word by word, modulo $2^{32}$, for $N=2^n$ words, using the bit-column/oblong machinery (a word column is the $(n+5)$-variate $\widetilde A$ with oblong view $\oA$ over a $5$-dimensional subspace $\D\subset\F_{2^{128}}$, $|\D|=32$). The prover commits the addends $\widetilde A,\widetilde B$ and the carry-out $\widetilde c_{\mathrm{out}}$; two virtual columns are the carry-in $\ocin=\ocout\ll1$ and the sum $\oC=\oA+\oB+\ocin$. The carry is correct exactly when
\begin{equation}
\oQ \;=\; \oA\cdot\oB \;+\; \oA\cdot\ocin \;+\; \oB\cdot\ocin \;+\; \ocout \;=\;0
\quad\text{on } \D\times\cube n,
\label{eq:Q}
\end{equation}
a single $\F_{2^{128}}$-identity of degree $\le2$ in each word variable and $\le62$ in $\II$. Its overflow $e$ is bit $31$ of $\widehat c_{\mathrm{out}}$. The zerocheck for \eqref{eq:Q} runs a univariate skip over the $32$ bit-positions (first round, degree $\le62$, reconstructed from $31$ values off $\D$ since $g$ vanishes on $\D$ when the statement holds) followed by $n$ ordinary degree-$\le3$ rounds; the final check needs $\oA,\oB,\ocout$ and $\ocin$ at $(r_{\II},\rho)$, obtained from one opening per committed column via the oblong reduction $\ocout(r_{\II},\rho)=\sum_{j}L_j(r_{\II})\,\widetilde c_{\mathrm{out}}(j,\rho)$ (the shift $\ocin$ relabels $j\mapsto j+1$). The full derivation and soundness ($O(n)/2^{128}$) are omitted here.

\section{Ring-switching for a Boolean witness (deferred)}\label{sec:ringswitch}

This appendix is \emph{unused by v1}, whose witness is field-valued and committed directly (\S\ref{sec:rs-dense}). It returns only with the bit-columns of Appendix~\ref{sec:add}: a \emph{Boolean} witness $q:\cube{M}\to\F_2$ committed with no embedding overhead, by packing $D=128$ bits per committed field element and bridging to a dense $\F_{2^{128}}$ PCS (DP24).

Pack each block of $D$ bits in the monomial basis, $g(y)=\sum_{w\in\cube\nu}x^{\mathrm{int}(w)}q(y,w)$ with $\nu=\log_2 D=7$ and $m'+\nu=M$; then $g$ takes full field values and the dense PCS commits it at zero overhead. To open $\widehat q(\mathbf r)=\alpha$, split $\mathbf r=(r_{\mathrm{hi}},r_{\mathrm{lo}})$, send the $D$ values $s_w=\widehat q(r_{\mathrm{hi}},w)$ with $\alpha=\sum_w\eq(r_{\mathrm{lo}},w)s_w$, descend to $\F_2$ to get $D$ claims $\widehat s_u=\sum_y A_u(y)\widehat g(y)$ (with $\eq(r_{\mathrm{hi}},y)=\sum_u A_u(y)x^u$), batch them with $\xi$, and run one sumcheck $\sum_y B(y)\widehat g(y)=\beta$ ending in a claim $\widehat g(\mathbf r')$ the dense PCS opens. The concrete PCS opens at an \emph{oblong} point: of the $\nu$ packed variables it fuses the low $k=6$ into one coordinate $\II$ over $\D$ through the Lagrange basis, $\widehat q(\II,r_{\mathrm{rest}})=\sum_{j}L_j(\II)\widehat q(j,r_{\mathrm{rest}})$, the same oblong form as the addition gadget's columns, so the batch's first round is taken as that skip. The full reduction and soundness ($O(M)/2^{128}$) are omitted here.

\fi % end deferred annex

\begin{thebibliography}{9}

\bibitem{m3}
Irreducible.
\newblock Multi-Multiset Matching (M3).
\newblock Binius documentation.
\newblock \url{https://www.binius.xyz/basics/binius-v0/m3/}

\bibitem{basefold}
Hadas Zeilberger, Binyi Chen, and Benjamin Fisch.
\newblock BaseFold: Efficient Field-Agnostic Polynomial Commitment Schemes from Foldable Codes.
\newblock Cryptology ePrint Archive, Paper 2023/1705, 2023.
\newblock \url{https://eprint.iacr.org/2023/1705}

\bibitem{cryptoeprint:2024/504}
Benjamin~E. Diamond and Jim Posen.
\newblock Polylogarithmic Proofs for Multilinears over Binary Towers.
\newblock Cryptology ePrint Archive, Paper 2024/504, 2024.
\newblock \url{https://eprint.iacr.org/2024/504}

\bibitem{cryptoeprint:2024/108}
Angus Gruen.
\newblock Some Improvements for the PIOP for ZeroCheck.
\newblock Cryptology ePrint Archive, Paper 2024/108, 2024.
\newblock \url{https://eprint.iacr.org/2024/108}

\bibitem{cryptoeprint:2025/1187}
Andrija Novakovic and Guillermo Angeris.
\newblock Ligerito: A Small and Concretely Fast Polynomial Commitment Scheme.
\newblock Cryptology ePrint Archive, Paper 2025/1187, 2025.
\newblock \url{https://eprint.iacr.org/2025/1187}

\bibitem{whir}
Gal Arnon, Alessandro Chiesa, Giacomo Fenzi, and Eylon Yogev.
\newblock WHIR: Reed--Solomon Proximity Testing with Super-Fast Verification.
\newblock Cryptology ePrint Archive, Paper 2024/1586, 2024.
\newblock \url{https://eprint.iacr.org/2024/1586}

\bibitem{flock}
Succinct Labs.
\newblock flock: a batched-R1CS proving system for hash circuits over binary towers.
\newblock \url{https://github.com/succinctlabs/flock}

\bibitem{leanvm-doc}
leanEthereum.
\newblock leanVM: a minimal zkVM (\texttt{misc/minimal\_zkVM.tex}, \S Range checks, \S Switch statements).
\newblock \url{https://github.com/leanEthereum/leanVM}

\end{thebibliography}

\end{document}
